# Sampling distribution for $\bar X - \bar Y$

For two independent normal random variables $X$ and $Y$, with distributions $X\sim N(\mu_X,\,\sigma_X)$ and $Y\sim N(\mu_Y,\,\sigma_Y)$ respectively, the distribution of $\bar X-\bar Y$ is,

$$
\bar X-\bar Y\sim N\Bigg(\mu_X-\mu_Y,\,\sqrt{\frac{\sigma_X^2}{n_X}+\frac{\sigma_y^2}{n_Y}}\Bigg)
$$

where $n_X$ is the size of the sample taken from the distribution for $X$ and $n_Y$ is the size of the sample taken from the distribution for $Y$.

We can show that this result is true by sampling from the distributions of $X$ and $Y$ in R and creating a new sampling distribution for $\bar X -\bar Y$.

Specifically, we will;

1. Draw $m=$ 20,000 samples of size $n_X=$ 120 from the distribution $X\sim N(90,\,12)$ and calculate the mean of each of these samples.
2. Draw $m=$ 20,000 samples of size $n_Y=$ 96 from the distribution $Y\sim N(40,\,10)$ and calculate the mean of each of these samples.
3. Calculate the difference between the sample means from the distribution for $X$ and from the distribution for $Y$ - that is $\bar X-\bar Y$.
4. Check whether the kernel density estimate of the distribution of these differences is similar to the expected normal distribution, as stated above.

::: {.webex-box .webex-check}
**Using the statement above about the distribution of $\bar X-\bar Y$, what distribution would you expect the differences in sample means to follow?**

```{r mcq-opts3, echo=FALSE}
opts3 <- sample(c(answer = "$N\\Big(50,\\,\\sqrt{\\textstyle\\frac{12^2}{120}+\\textstyle\\frac{10^2}{96}}\\Big)$",
           "$N\\Big(50,\\,\\sqrt{\\textstyle\\frac{12}{120}+\\textstyle\\frac{10}{96}}\\Big)$",
           "$N\\Big(50,\\,\\sqrt{\\textstyle\\frac{12^2}{20,000}+\\textstyle\\frac{10^2}{20,000}}\\Big)$",
           "$N\\Big(90,\\,\\sqrt{\\textstyle\\frac{12^2}{120}+\\textstyle\\frac{10^2}{96}}\\Big)$"), size = 4)
```

`r longmcq(opts3)`

`r hide("Solution")`
We know that $X\sim N(90,\,12)$ and that we want to take samples of size 120 from this distribution. Therefore, we can say that $\mu_X=90$, $\sigma_X=12$ and $n_X=120$.

We also know that $Y\sim N(40,\,10)$ and that we want to take samples of size 96 from this distribution, so $\mu_Y=40$, $\sigma_Y=10$ and $n_Y=96$.

Plugging these values into the given distribution for $\bar X-\bar Y$ tells us that,

$$
\bar X-\bar Y\sim N\Big(90-40=50,\,\sqrt{\textstyle\frac{12^2}{120}+\textstyle\frac{10^2}{96}}\Big)
$$
`r unhide()`
:::

::: {.webex-box .webex-check}
**Complete the code below draw 20,000 samples of size 120 from the $N(90,\, 12)$ distribution and save the mean of each sample in a vector called `meansX`.**

**Let `m` represent the number of samples you want to draw, and `nx` represent the size each sample should be i.e. $n_X$.**

::: {.webex-code}
`m <- ``r fitb(20000)`

<ul style="margin-top: -0.8em; margin-left:-2em;">
`nx <- ``r fitb(120)`
</ul>
<ul style="margin-top:1.6em; margin-left:-2em;">
`meansX <- ``r mcq(c("vector(m)", answer="numeric(m)", "c(m)", "list(m)", "data.frame(m)"))`
</ul>

<ul style="margin-top:1.6em; margin-left:-2em;">
`r mcq(c(answer="for", "while", "if"))``(``r fitb(c("i in 1:m", "i in 1:20000"))``){`
</ul>
<ul style="margin-top:-0.8em; margin-left:0.6em;">
`meansX[i] <- mean(``r fitb("rnorm")``(nx, mean = ``r fitb(90)``, sd = ``r fitb(12)``))`
</ul>
<ul style="margin-top:-0.8em; margin-left:-2em;">
`}`
</ul>

:::

`r hide("Solution")`
```{r XY-ans1}
m <- 20000
nx <- 120

meansX <- numeric(m)

for(i in 1:m){
  meansX[i] <- mean(rnorm(nx, mean = 90, sd = 12))
}
```
`r unhide()`
:::

Similarly, we can draw 20,000 samples of size $n_Y=$ 96 from the $N(40,\, 10)$ distribution and calculate the mean of each sample using the following code. As we've seen before, this makes use of a for loop to draw each sample and calculate its mean, before saving the mean in the vector `meansY`.

```{r XY1}
m <- 20000
ny <- 96

meansY <- numeric(m)

for(i in 1:m){
  meansY[i] <- mean(rnorm(ny, mean = 40, sd = 10))
}
```

We can then calculate the difference between each pair of means and store these differences in the data frame `XY`. We use a data frame, rather than a vector here, so that we can go on to plot the distribution of differences using `ggplot2`.

```{r XY2}
XY <- data.frame(diff = meansX - meansY)
```



Finally, we can use `ggplot2` to plot the kernel density estimate of the differences between each pair of sample means from the $N(90,\,12)$ and $N(40,\,10)$ distributions. We do this using the `geom_density()` function.

We know that $\bar X-\bar Y$ follows the $N\Big(50,\,\sqrt{\textstyle\frac{12^2}{120}+\textstyle\frac{10^2}{96}}\Big)$ distribution, so we can superimpose this above the kernel density estimate to assess how similar they are. We do this by including the layer `stat_function()`. Because the distribution we want to show is a normal distribution, we use the argument `fun = dnorm`. The mean and standard deviation, $50$ and $\sqrt{\textstyle\frac{12^2}{120}+\textstyle\frac{10^2}{96}}$ respectively, are included within a list given to the argument `args =`. We can also set some stylistic choices of how this normal distribution should appear using the `col =`, `linetype =`, `fill =`, `alpha =` and `geom =` arguments.

```{r XY3}
ggplot(data = XY) +
  geom_density(aes(x = diff), fill = "cadetblue3") +
  stat_function(fun = dnorm, args = list(mean = 50, sd = sqrt(12^2/120 + 10^2/96)),
                geom = "area", col = "black", linetype = 2, fill = "black", alpha = 0.2) +
  labs(x = expression(bar(X)-bar(Y)), y = "Density")
  
```

::: {.webex-box .webex-check}
**Do you think the kernel density estimate of the distribution of the differences between pairs of sample means is approximately equal to the $N\Big(50,\,\sqrt{\textstyle\frac{12^2}{120}+\textstyle\frac{10^2}{96}}\Big)$ distribution?**

`r mcq(c(answer="Yes", "No", "Can't say"))`
:::

<!--computing empirical and theoretical probability-->
We can use our calculated differences between the sample means of $X$ and $Y$ to estimate probabilities for $\bar X-\bar Y$. For example, you might be asked, what is the probability that the difference between the mean of $X$ and the mean of $Y$ is greater than 48?

We can use the values of differences between the pairs of sample means to answer this. We use a logical statement, `XY$diff > 48`, to identify which of the calculated differences are greater than 48. This logical statement returns 20,000 values - `TRUE` if the corresponding difference was greater than 48 and `FALSE` if the difference was less than or equal to 48. We can then plug this logical vector into the `mean()` function to tell us the proportion of differences that were greater than 48. This proportion is a good approximation of the true probability $P\big(\bar X-\bar Y>48\big)$ and using the code below, we see that this is roughly equal to `r mean(XY$diff > 48)`.

```{r XY4}
mean(XY$diff > 48)
```

Because we have used our samples to calculate this probability, this is known as an *empirical* probability. We could also find the true, theoretical probability that $P\big(\bar X-\bar Y>48\big)$ by using the fact that $\bar X-\bar Y\sim N\Big(50,\,\sqrt{\textstyle\frac{12^2}{120}+\textstyle\frac{10^2}{96}}\Big)$.

We can use the function `pnorm()` to find the probability that $P\big(\bar X-\bar Y\leq 48\big)$. Then, in order to find $P\big(\bar X-\bar Y>48\big)$, we simply calculate $1-$ the result from `pnorm()`. This is shown in the code below.

```{r XY5}
1-pnorm(48, mean = 50, sd = sqrt(12^2/120 + 10^2/96))
```

Notice how close the empirical probability we calculated from our samples is to the true, theoretical probability.

---

You can follow a similar example of using R to create a sample sampling distribution for $\bar X-\bar Y$ in Section [6.5.2 Sampling Distribution for $\bar X-\bar Y$ When Sampling from Two Independent Normal Populations](https://read.kortext.com/reader/pdf/92741/375) in *Probability and Statistics with R*.









