# Cluster Sampling

::: {.webex-box}
**Example** Suppose a company that gives city tours wants to survey its customers. They ask every customer to rate their experience on a scale of 1 to 10. One tour guide runs ten tours a day, from which four should be randomly selected and all the customer ratings from these should be included in the sample.

Randomly select these clusters of customers for the company and include their ratings in the sample.

`r hide("Solution")`
To start, set up a data frame to contain the customer ratings and the tour they belonged to. Each tour is given to 20 customers.

```{r}
tours <- data.frame(tour = rep(1:10, each = 20),
                    rating = sample(x = 1:10, size = 200, replace = TRUE,
                                        prob = c(rep(1/20, 6), rep(1/5, 3), 1/10)))
head(tours)
```

In order to randomly select the tours to use as the clusters, use the `sample()` function.

```{r}
clusters <- sample(1:10, size = 4, replace = FALSE)
clusters
```

Then, the sample includes all customers who were on a tours `r sort(clusters)[1:3]` or `r sort(clusters)[4]`.
```{r}
cluster_sample <- tours[tours$tour %in% clusters, ]
```

We can verify that all customers from each tour group chosen as a cluster are included in the sample.
```{r}
table(cluster_sample$tour)
```
`r unhide()`
:::